`methyvim`: Targeted and Model-free Differential Methylation Analysis in

### 2018-06-22 paper/paper_BiocF1000.Rmd
integrate RVP revisions/comments

---
title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
  R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing a general algorithm for the
  nonparametric estimation of treatment effects on DNA methylation at CpG sites
  throughout the genome, complete with straightforward statistical inference for
  such estimates. The approach leverages variable importance measures derived
  from statistical parameters arising causal inference, defined in such a manner
  that they may be used to obtain targeted estimates of the relative importance
  of individual CpG site with respect to a binary treatment assigned at the
  phenotype level. The procedure implemented is computationally efficient,
  incorporating a preliminary screening step to isolate a subset of sites for
  which there is cursory evidence of differential methylation as well as a
  unique multiple testing correction to control the False Discovery Rate with
  the same rigor as if all sites were tested. This technique for analysis of
  differentially methylated positions provides an avenue to incorporate flexible
  state-of-the-art machine learning algorithms into the estimation of
  differential methylation effects without the loss of interpretable statistical
  inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, targeted minimum loss-based
  estimation, machine learning
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA methylation occurs at CpG sites
and involves the addition of a methyl group ($\text{CH}_3$) to the fifth carbon
of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
medical studies have implicated DNA methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to rigorously probe the molecular mechanisms of this
epigenetic process. Modern assays, like the Illumina _Infinium_ Methylation chip
technology, allow for quantitative interrogation of DNA methylation at a large
set of CpG sites scattered across the genome at single-nucleotide resolution;
moreover, much effort has been invested, by the bioinformatics community, in the
development of tools for properly removing technological effects that may
contaminate biological signatures measured by such assays
[@fortin2014functional, dedeurwaerder2013comprehensive]. Despite these
advances in both biological and bioninformatical techniques, most statistical
methods available for differential analysis of data produced by such assays rely
on over-simplified models that fail to incorporate many of the recent advances
in machine learning. Generally, typical statistical estimation approaches do not
readily extend to such high-dimensional data without restrictive modeling
assumptions and the use of inferentially costly hypothesis testing corrections.
When these standard assumptions are violated, estimates of the population-level
effect of an exposure or treatment may suffer from large bias. Data-adaptive
estimation procedures that utilize machine learning can control for confounding
even in high-dimensional settings; however, interpretable statistical inference
(i.e., confidence intervals and hypothesis tests) resulting from these
data-adaptive estimates is challenging to obtain [@libbrecht2015machine].

In this paper, we briefly present an alternative to such statistical analysis
approaches in the form of a nonparametric estimation procedure that provides
simple and readily interpretable statistical inference, discussing at length a
recent implementation of the methodology in the `methyvim` R package. Inspired
by recent advances in statistical causal inference and machine learning, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric _variable importance measures_ (VIMs) [@vdl2006statistical],
estimated at a set of pre-screened CpG sites, controlling for the False
Discovery Rate (FDR) as if all sites were tested. Under standard assumptions
(e.g., identifiability, strong ignorability) [@pearl2009causality], targeted
minimum loss-based estimators of regular asymptotically linear estimators have
sampling distributions that are asymptotically normal, allowing for reliable
point estimation and the construction of Wald-style confidence intervals
[@vdl2011targeted, @vdl2018targeted]. As counterfactuals are contrary-to-fact --
that is, defined in terms of missing data -- in the context of DNA methylation,
we define the counterfactual outcomes under a binary treatment as the observed
methylation (whether Beta- or M-) value a CpG site would have if all subjects
were given the treatment and the methylation value a CpG site would have if
treatment were withheld from all subjects. Although these counterfactual
outcomes are, of course, impossible to observe, they have statistical analogs
that may be reliably estimated from observed data [@pearl2009causality]. We
describe an algorithm that incorporates the estimation of VIMs using _targeted
minimum loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
description and analysis of the statistical methodology to work outside the
scope of the present manuscript. This methodology assesses the individual
importance of a given CpG site by utilizing state-of-the-art machine learning
algorithms in the estimation of a targeted VIM, initially proposed and explored
in [@tuglus2008targeted]. In the presnet work, we focus on a software package,
`methyvim`, that implements a variant of this methodology, specifically tailored
to differential methylation analysis.

@@

We consider an observed data structure $O = (W_k, A, Y_j)$, where $Y_j, j = 1,
\ldots J$ represents a given CpG site of interest, $A \in \{0, 1\}$ represents a
binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
the observed values of CpG sites in the same neighborhood as $j$. We consider
the case of observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
take $O \sim \mathcal{P} \in \mathcal{M}$, which is to say that $O$ is governed
by an unknown probability distribution $\mathcal{P}$, taken only to reside in a
nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
data-generating process. We consider having access to measurements on a large
number $G$ of CpG sites (e.g., 850,000, as measured by the Illumina
MethylationEPIC BeadChip arrays), and as a matter of practicality, let $J \leq
G$, so that $J$ indexes only those CpG sites that pass an arbitrary
pre-screening procedure; note that the equality is only obtained in the case
that all sites pass the pre-screening procedure. Further, we let $K(j)$ index
the set of all CpG sites so that $k$, the realization of $K(j)$ for a given CpG
site $j$, is a set of the CpG sites that neighbor the CpG site $j$, where the
definition of a neighborhood is left as user-specified; thus, $W_k$ is a matrix
of observed methylation values for the set of CpG sites $k$ that neighbor a
particular target site $j$. We consider estimating a target parameter ($\psi_j$)
as a variable importance measure of the effect of changing a treatment $A$ on
the methylation of a CpG site $j$, accounting for the observed methylation at
the set of sites $k$ that neighbor the site $j$. After a complete run of the
procedure, we have access to $j$ VIM estimates $\psi_j$, each corresponding to
the degree to which a given CpG site $j$ was responsive to a change in the
binary treatment $A$, after controlling for any confounding exerted by its
neighboring CpG sites. When machine learning estimators are employed in
estimating $\psi_j$, this produces a nonparametric variable importance measure
of differential methylation, allowing for the identification of differentially
methylated positions (DMPs).

@@

* `tmle.npvi` - A nonparametric variable importance measure (NPVI)
  [@chambaz2012estimation]. The effect of a continuous-valued exposure or
  treatment (the observed methylation at a target CpG site) on an outcome of
  interest is estimated, controlling for the observed methylation value at all
  other CpG sites in the same neighborhood as the target site. This uses a
  parameter that compares values of the treatment against a user-specified
  reference value taken to be the null. In particular, the implementation to be
  provided is designed to assess the effect of differential methylation at the
  target CpG site on an outcome of interest (e.g., survival), providing a
  nonparametric evaluation of the impact of methylation at the target site.

* `txshift` - Variable importance using a causal parameter defined as the
  counterfactual outcome under a posited shift of a continuous-value treatment
  of interest using stochastic intervention policies [@munoz2012population,
  @diaz2018stochastic, @hejazi2018txshift]. The value of an outcome of interest
  under an unobserved value of the (continuous-valued) treatment, specified
  through a user-provided _additive shift_ of the observed treatment, may be
  data-adaptively estimated. This allows for the effect of changes/shifts in the
  observed methylation at a CpG site on an outcome of interest to be evalauted,
  providing a nonparametric evalution of the relative importance of a particular
  target CpG site with respect to another variable measured in the same study.

@@

```{r methyvim-heatmap}
methyheat(methyvim_cancer_ate)
```

Invoking `methyheat` in this manner produces a plot of the top sites ($25$, by
default) based on the raw p-value, using the raw methylation measures in the
plot. This uses the exceptional `superheat` R package [@barter2017superheat].

>>>>

---
title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
  R'
author:
  - name: Nima S. Hejazi
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Center for Computational Biology, University of California, Berkeley
    orcid: 0000-0002-7127-2789
    email: nhejazi@berkeley.edu
  - name: Rachael V. Phillips
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation:
      - Group in Biostatistics, University of California, Berkeley
      - Department of Statistics, University of California, Berkeley
    orcid: 0000-0003-1432-5511
abstract: |
  We present `methyvim`, an R package implementing a general algorithm for the
  nonparametric estimation of treatment effects on DNA methylation at CpG sites
  throughout the genome, complete with straightforward statistical inference for
  such estimates. The approach leverages variable importance measures derived
  from statistical parameters arising causal inference, defined in such a manner
  that they may be used to obtain targeted estimates of the relative importance
  of individual CpG sites with respect to a binary treatment assigned at the
  phenotype level. The procedure implemented is computationally efficient,
  incorporating a preliminary screening step to isolate a subset of sites for
  which there is cursory evidence of differential methylation as well as a
  unique multiple testing correction to control the False Discovery Rate with
  the same rigor as if all sites were tested. This technique for analysis of
  differentially methylated positions provides an avenue to incorporate flexible
  state-of-the-art machine learning algorithms into the estimation of
  differential methylation effects without the loss of interpretable statistical
  inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, machine learning, targeted minimum
  loss-based estimation
bibliography: references.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA methylation mostly commonly
occurs at CpG sites and involves the addition of a methyl group ($\text{CH}_3$)
to the fifth carbon of the cytosine ring structure to form 5-methylcytosine.
Numerous biological and medical studies have implicated DNA methylation as
playing a role in disease and development [@robertson2005dna]. Perhaps
unsurprisingly then, biotechnologies have been developed to rigorously probe the
molecular mechanisms of this epigenetic process. Modern assays, like the
Illumina _Infinium_ HumanMethylation BeadChip assay, allow for quantitative
interrogation of DNA methylation, at single-nucleotide resolution, across a
comprehensive set of CpG sites scattered across the genome; moreover, the
computational biology community has invested significant effort in the
development of tools for properly removing technological effects that may
contaminate biological signatures measured by such assays
[@fortin2014functional, @dedeurwaerder2013comprehensive]. Despite these
advances in both biological and bioninformatical techniques, most statistical
methods available for differential analysis of data produced by such assays rely
on over-simplified models that do not readily extend to such high-dimensional
data structures without restrictive modeling assumptions and the use of
inferentially costly hypothesis testing corrections. When these standard
assumptions are violated, estimates of the population-level effect of an
exposure or treatment may suffer from large bias. What's more, reliance on
restrictive and misspecified statistical models naturally leads to biased effect
estimates that are not only misleading in assessing effect sizes but also result
in false discoveries as these biased estimates are subject to testing and
inferential procedures. Such predictably unreliable methods serve only to
produce findings that are later invalidated by replication studies and add still
further complexity to discovering biological targets for potential therapeutics.
Data-adaptive estimation procedures that utilize machine learning provide a way
to overcome many of the problems common in classical methods, controlling for
potential confounding even in high-dimensional settings; however, interpretable
statistical inference (i.e., confidence intervals and hypothesis tests) from
such data-adaptive estimates is challenging to obtain [@libbrecht2015machine].

In this paper, we briefly present an alternative to such statistical analysis
approaches in the form of a nonparametric estimation procedure that provides
simple and readily interpretable statistical inference, discussing at length a
recent implementation of the methodology in the `methyvim` R package. Inspired
by recent advances in statistical causal inference and machine learning, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric _variable importance measures_ (VIMs) [@vdl2006statistical],
estimated at a set of pre-screened CpG sites, controlling for the False
Discovery Rate (FDR) as if all sites were tested. Under standard assumptions
(e.g., identifiability, strong ignorability) [@pearl2009causality], targeted
minimum loss-based estimators of regular asymptotically linear estimators have
sampling distributions that are asymptotically normal, allowing for reliable
point estimation and the construction of Wald-style confidence intervals
[@vdl2011targeted, @vdl2018targeted]. As counterfactuals are contrary-to-fact --
that is, defined in terms of missing data -- in the context of DNA methylation,
we define the counterfactual outcomes under a binary treatment as the observed
methylation (whether Beta- or M-) values a CpG site would have if all subjects
were given the treatment and the methylation values a CpG site would have if
treatment were withheld from all subjects. Although these counterfactual
outcomes are, of course, impossible to observe, they have statistical analogs
that may be reliably estimated from observed data [@pearl2009causality]. We
describe an algorithm that incorporates the estimation of VIMs using _targeted
minimum loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
description and analysis of the statistical methodology to work outside the
scope of the present manuscript. This methodology assesses the individual
importance of a given CpG site by utilizing state-of-the-art machine learning
algorithms in the estimation of a targeted VIM, initially proposed and explored
in [@tuglus2008targeted]. In the present work, we focus on a software package,
`methyvim`, that implements a variant of this methodology, specifically tailored
to differential methylation analysis.

@@

We consider an observed data structure $O = (W_k, A, Y_j)$, where $Y_j, j = 1,
\ldots J$ represents a given CpG site of interest, $A \in \{0, 1\}$ represents a
binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
the observed values of CpG sites in the same neighborhood as $j$ as well as any
potential phenotype-level confounders (e.g., sex, age). We consider having
access to measurements on a large number $G$ of CpG sites (e.g., $850,000$, as
measured by the Illumina MethylationEPIC BeadChip arrays), and as a matter of
practicality, let $J \leq G$, so that $J$ indexes only those CpG sites that pass
an arbitrary pre-screening procedure; note that equality is only obtained in the
case that all sites pass the pre-screening procedure. Further, we let $K(j)$
index the set of all CpG sites so that $k$, the realization of $K(j)$ for a
given CpG site $j$, is a set of the CpG sites that neighbor the CpG site $j$,
where the definition of a neighborhood is left as user-specified; thus, $W_k$ is
a matrix of observed methylation values for the set of CpG sites $k$ that
neighbor a particular target site $j$ as well as any potential phenotype-level
confounders, where the latter does not vary between realizations of $K(j)$. As
a variable importance measure, we see to estimate a target parameter ($\psi_j$)
that quantifies the effect of changing a treatment $A$ on the methylation of a
CpG site $j$, accounting for any potential confounding from phenotype-level
covariates and the observed methylation at the set of sites $k$ that neighbor
the site $j$. After a complete run of the procedure, we have access to $j$ VIM
estimates $\psi_j$, each corresponding to the degree to which a given CpG site
$j$ was responsive to a change in the binary treatment $A$, after controlling
for potential confounding. When machine learning estimators are employed in
estimating $\psi_j$, this produces a nonparametric variable importance measure
of differential methylation, allowing for the identification of differentially
methylated positions (DMPs). We note that $\psi(j)$, for any given $j$, is a
function of the _true_ data-generating distribution; so, to formalize, we
consider observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
have $O \sim P \in \mathcal{M}$, which is to say that the random variable $O$ is
governed by an unknown probability distribution $P$, assumed only to reside in a
nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
data-generating process.

@@

* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]
  (R package `tmle.npvi`). The effect of a continuous-valued exposure or
  treatment (the observed methylation at a target CpG site) on an outcome of
  interest is estimated, controlling for the observed methylation value at all
  other CpG sites in the same neighborhood as the target site. This uses a
  parameter that compares values of the treatment against a user-specified
  reference value taken to be the null value. In particular, the implementation
  to be provided is designed to assess the effect of differential methylation at
  the target CpG site on an outcome of interest (e.g., survival), providing a
  nonparametric evaluation of the impact of methylation at the target site.

* The causal effect of shifting the value of an observed intervention, defined
  as the counterfactual outcome under a posited shift of a continuous-value
  treatment of interest using stochastic intervention policies
  [@munoz2012population, @diaz2018stochastic, @hejazi2018txshift] (R package
  `txshift`). The value of an outcome of interest under an unobserved value of
  the (continuous-valued) treatment, specified through a user-provided
  _additive shift_ of the observed treatment, may be data-adaptively estimated.
  This allows for the effect of changes/shifts in the observed methylation at a
  CpG site on an outcome of interest to be evaluated, providing a nonparametric
  evalution of the relative importance of a particular target CpG site with
  respect to another variable measured in the same study.

@@

```{r methyvim-heatmap}
methyheat(methyvim_cancer_ate, smooth.heat = TRUE, left.label = "none")
```

__Remark__: Invoking `methyheat` in this manner produces a plot of the top sites
($25$, by default) based on the raw p-value, using the raw methylation measures
in the plot. This uses the exceptional `superheat` R package
[@barter2017superheat], to which we can easily pass additional parameters. In
particulat, we hide the CpG site labels that would appear by default on the left
of the heatmap (by setting `left.label = "none"`) to emphasize that this is only
an example and _not_ a scientific discovery.
